Trends and Challenges in Toronto’s Homeless Population (2018-2024)

### 2024-09-27 paper/paper.qmd
Final commit

Homelessness is a growing global concern, especially with rapid urbanization and the housing crisis. As Canada's largest city, Toronto faces a growing homeless population [@toronto2024]. In addition to being a socio-economic issue, homelessness intersects with public health and social justice [@healthhomelessness]. Understanding the trends and composition of the homeless population is critical to effective policymaking. However, existing data often lack detailed categorization, hindering a full understanding of the needs of specific groups. This study seeks to fill this gap.

Using the "Toronto Shelter System Flows" dataset form Toronto Open Data Platform [@torontosheltersystem], this paper examines trends in Toronto's homeless population from 2018 to 2024, focusing on age, gender, and chronic homelessness. Our analysis reveals a decline during the pandemic, followed by a rebound, with males and the 35-44 age group comprising the majority. The number of chronically homeless people continues to grow, providing important insights for policymakers.

In this essay, @sec-data-overview explains the background and necessity of using this dataset, followed by an explanation of the variables mentioned and the methods used. @sec-data-results shows the changes in the number of homeless individuals, along with the proportions and trends over time for each age group, gender, and chronic homelessness. Graphs and related explanations are provided. @sec-discussion further analyses the information provided in the figures and describes the weaknesses of the study.

@@

The dataset used for this analysis is the “Toronto Shelter System Flows” dataset from the Toronto Open Data Platform [@torontosheltersystem]. It focuses on the measurement of homeless individuals' flow within Toronto’s shelter system, which is funded and operated by the City of Toronto. It covers data from 2018 to September 2024, with the homeless count based on shelter admission records, including demographic information such as age and gender. The dataset also provides data on specific groups, including chronic homeless individuals, refugees, youth, single adults, and families. The measurements are collected through the Shelter Management Information System (SMIS) and are updated monthly to ensure that trends in the homeless population are accurately and timely reflected [@torontosheltersystem].

@@

@fig-gender-distribution-sidebyside shows the gender distribution of Toronto's homeless population from 2018 to 2024 and how it has changed over time. Males consistently represent the largest proportion, making up approximately 61.8% of the total homeless population, while females account for about 36.8%, and transgender or non-binary individuals make up around 1.4%. The percentages of males and transgender or non-binary individuals initially increase but begin to decline after 2021, with the opposite trend observed for females. However, these fluctuations are minor, staying within five percentage points.

@@

@fig-chronic-population displays the changes in the number and proportion of chronic and non-chronic homeless people in Toronto between 2018 and 2024. The number of chronically homeless has been growing over this time, with over 30,000 people in 2018, and by 2023, the number has gone to about 60,000, nearly doubling. The slight drop in numbers in 2024 is due to the fact that the statistics were last updated on September 17, 2024, and it is projected that by the end of the year, the number will have exceeded the 2023 one. Similarly, the percentage of chronically homeless is growing every year, from 29.7% in 2019 to 63.7% in 2024. Note that in 2019 to 2020 and 2023 to 2024, the rate grows faster, by about 10 percent.

@@

Analysing age, gender and chronic homelessness helps to understand Toronto's homeless population. Homelessness is concentrated among working-age adults, particularly those aged 35-44 and 25-34, likely due to challenges with employment or housing stability. The sharp decline in the under-16 group during the pandemic may reflect government support for youth, but the subsequent increase underscores the need for sustained policies. In terms of gender, men remain the majority, with only 1.4% of people being transgender or non-binary. This suggests that shelters should design facilities to meet the different needs of men and women, while also accommodating transgender or non-binary people. The increase in the number of chronically homeless people highlights the inadequacy of current interventions to help people escape homelessness in the long term, and signals a worsening situation without further intervention.

>>>>

Homelessness is a growing global concern, especially with rapid urbanization and the housing crisis. As Canada's largest city, Toronto faces a growing homeless population [@toronto2024]. In addition to being a socio-economic issue, homelessness intersects with public health and social justice [@healthhomelessness].  However, the existing data often lacks detailed classification, preventing a comprehensive understanding of the needs of specific groups. The present study seeks to fill this gap by analyzing the share and trends of different groups in the homeless population.

Using the "Toronto Shelter System Flows" dataset from the Toronto Open Data Platform [@torontosheltersystem], this paper examines trends in Toronto's homeless population from 2018 to 2024, focusing on age, gender, and chronic homelessness. Our analyses show that homelessness declined during the pandemic and then rebounded. Among all homeless people, males and the 35-44 age group make up the majority, and the number of chronically homeless people continues to grow. These results provide important insights for policymakers.

In this essay, @sec-data-overview explains the background and necessity for using the dataset, followed by an explanation of the dataset's measurements. @sec-data-results shows the changes in the number of homeless individuals, along with the proportions and trends over time for each age group, gender, and chronic homelessness. Graphs and related explanations are provided. @sec-discussion further analyses the information provided in the figures and describes the weaknesses of the study.

@@

The dataset used for this analysis is the “Toronto Shelter System Flows” dataset from the Toronto Open Data Platform [@torontosheltersystem]. It focuses on the measurement of homeless individuals' flow within Toronto’s shelter system, which is funded and operated by the City of Toronto. It covers data from 2018 to September 2024, with the homeless count based on shelter admission records, including demographic information such as age and gender. The dataset also provides data on specific groups, including chronically homeless individuals, refugees, youth, single adults, and families. The measurements are collected through the Shelter Management Information System (SMIS) and are updated monthly to ensure that trends in the homeless population are accurately and timely reflected [@torontosheltersystem].

@@

@fig-gender-distribution-sidebyside shows the gender distribution of Toronto's homeless population from 2018 to 2024 and how it has changed over time. Males consistently represent the largest proportion, making up approximately 61.8% of the total homeless population, while females account for about 36.8%, and transgender or non-binary individuals makeup around 1.4%. The percentages of males and transgender or non-binary individuals initially increase but begin to decline after 2021, with the opposite trend observed for females. However, these fluctuations are minor, staying within five percentage points.

@@

@fig-chronic-population displays the changes in the number and proportion of chronic and non-chronic homeless people in Toronto between 2018 and 2024. The number of chronically homeless has been growing over this time, with over 30,000 people in 2018, and by 2023, the number has gone to about 60,000, nearly doubling. The slight drop in numbers in 2024 is due to the fact that the statistics were last updated on September 17, 2024, and it is projected that by the end of the year, the number will have exceeded the 2023 one. Similarly, the percentage of chronically homeless is growing every year, from 29.7% in 2019 to 63.7% in 2024. It should be noted that this percentage grows faster at about 10 percent between 2019 to 2020 and 2023 to 2024.

@@

Analyzing age, gender and chronic homelessness helps to understand Toronto's homeless population. Homelessness is concentrated among working-age adults, particularly those aged 35-44 and 25-34, likely due to challenges with employment or housing stability. The sharp decline in the under-16 group during the pandemic may reflect government support for youth, but the subsequent increase underscores the need for sustained policies. In terms of gender, men remain the majority, with only 1.4% of people being transgender or non-binary. This suggests that shelters should design facilities to meet the different needs of men and women, while also accommodating transgender or non-binary people. The increase in the number of chronically homeless people highlights the inadequacy of current interventions to help people escape homelessness in the long term and signals a worsening situation without further intervention.
